\input pdfToolbox
\input syntax-hi
\input preamble

\setlayout{horizontal margin=2cm, vertical margin=2cm}

{\setbox0=\hbox{\setfontandscale{bf}{25pt}Linear Expansion}
\centerline{
    \vbox{
        \copy0
        \smallskip
        %\hbox to\wd0{\hfil Ari Feiglin}
    }
}}

\bigskip
\hbox to\hsize{\hfil\vbox{\hsize=.7\hsize
\leftskip=0pt plus 1fil \rightskip=\leftskip \parfillskip=\z@
\hrule
\kern5pt

In this paper I will define the concept of linear expansion in the context of syntax parsing.
We will progress through more and more complicated examples, beginning from the programming of a simple calculator until we ultimately have created an extensible programming language.

\kern5pt
\hrule
}\hfil}

\section*{Table of Contents}

\tableofcontents

\vfill\break

\section{Theoretical Background}

\subsection{Stateless Expansion}

The idea of linear expansion is simple: given a string $\xi$ the first character looks if it can bind with the second character to produce a new character, and the process repeats itself.
There is of course, nuance.
This nuance hides in the statement ``if it can bind'': we must define the rules for binding.

Let us define an {\it expander} to be a tuple $(\Sigma,\beta)$ where $\Sigma$ is an alphabet and $\beta\colon\overline\Sigma\times\overline\Sigma\plongto\overline\Sigma$ is a partial function called the
{\it reduction function} where $\overline\Sigma=\Sigma\times\overline{\bb N}$, $\overline{\bb N}\coloneqq{\bb N}\cup\set\infty$.
A {\it program} over an expander is a string over $\overline\Sigma$.
We write a program like $\sigma^1_{i_1}\cdots\sigma^n_{i_n}$ instead of as pairs $(\sigma^1,i_1)\dots(\sigma^n,i_n)$.
In the character $\sigma_i$, we call $i$ the {\it priority} of $\sigma$.

Then the rules of reduction are as follows, meaning we define $\beta(\xi)$ for a program:
We do so in cases:
\benum
    \item If $\xi=\sigma_i$ then $\beta(\sigma)=\sigma_0$.
    \item If $\xi=\sigma^1_i\sigma^2_j\xi'$ where $i\geq j$ and $\beta(\sigma^1,\sigma^2)=\sigma^3_k$ is defined then $\beta(\xi)=\sigma^3_k\xi'$.
    \item Otherwise, $\beta(\xi)=\sigma^1_i\beta(\sigma^2_j\xi')$.
\eenum

A string $\xi$ such that $\beta(\xi)=\xi$ is called {\it irreducible}.
Notice that it is possible for a string of length more than $1$ to be irreducible: for example if $\beta(\sigma^1,\sigma^2)$ is not defined then $\sigma^1_i\sigma^2_j$ is irreducible.
$$ \beta(\sigma_1\tau_2)\xvarrightarrow{(3)}\sigma_1\beta(\tau_2)\xvarrightarrow{(1)}\sigma_1\tau_2 $$
But such strings are not desired, since in the end we'd like a string to give us a value.
So an irreducible string which is not a single character is called {\it ill-written}, and a string which is not ill-written is {\it well-written}.

{\bf Example:\/} let $\Sigma={\bb N}\cup\set{+,\cdot}\cup\set{(n+),(n\cdot)}[n\in{\bb N}]$.
$\beta$ as follows:

\medskip
\centerline{
    \vtop{\ialign{\hfil$#$\hfil\tabskip=.25cm&\hfil$#$\hfil\cr
        \sigma^1_i,\sigma^2_j & \beta(\sigma^1_i,\sigma^2_j)\cr\noalign{\kern3pt\hrule\kern3pt}
        n,+ & (n+)\cr
        n,\cdot & (n\cdot)\cr
        (n+),m & n+m\cr
        (n\cdot),m & n\cdot m\cr
        (n\cdot),(m+) & (n\cdot m,+)\cr
        (n+),(m+) & (n+m,+)\cr
        (n\cdot),(m\cdot) & (n\cdot m,\cdot)\cr
    }}
}
\medskip
\noindent Where $n,m$ range over all values in ${\bb N}$.
Here $\beta(\sigma_i,\sigma_j)$'s priority is $j$.

Now let us look at the string $1+2\cdot3+4;$.
Here,
$$ \eqalign{
    1_\infty+_12_\infty\cdot_23_\infty+_14_\infty &\longto (1+)_12_\infty\cdot_23_\infty+_14_\infty\cr
    &\longto (1+)_1(2\cdot)_23_\infty+_14_\infty\cr
    &\longto(1+)_1(2\cdot)_2(3+)_14_\infty\cr
    &\longto(1+)_1(6+)_14_\infty\cr
    &\longto(7+)_14_\infty\cr
    &\longto(7+)_14_0\cr
    &\longto(11)_0
} $$
So the rules for $\beta$ we supplied seem to be sufficient for computing arithmetic expressions following the order of operations.

{\bf Example:\/} We can also expand our language to include parentheses.
So our alphabet becomes $\Sigma={\bb N}\cup\set{+,\cdot,(,)}\cup\set{\bop n+\bcp,\bop n\cdot\bcp,\bop n)\bcp}[n\in{\bb N}]$.
We distinguish between parentheses and bold parentheses for readability.
We extend $\beta$ as follows:

\medskip
\centerline{
    \vtop{\ialign{\hfil$#$\hfil\tabskip=.25cm&\hfil$#$\hfil\cr
        \sigma^1_i,\sigma^2_j & \beta(\sigma^1_i,\sigma^2_j)\cr\noalign{\kern3pt\hrule\kern3pt}
        n,+ & \bop n+\bcp_j\cr
        n,\cdot & \bop n\cdot\bcp_j\cr
        \bop n+\bcp,m & (n+m)_j\cr
        \bop n\cdot\bcp,m & (n\cdot m)_j\cr
        \bop n\cdot\bcp,\bop m+\bcp & \bop n\cdot m,+\bcp_j\cr
        \bop n+\bcp,\bop m+\bcp & \bop n+m,+\bcp_j\cr
        \bop n\cdot\bcp,\bop m\cdot\bcp & \bop n\cdot m,\cdot\bcp_j\cr
        n,) & \bop n)\bcp_j\cr
        \bop n+\bcp,\bop m)\bcp & \bop n+m)\bcp_j\cr
        \bop n\cdot\bcp,\bop m)\bcp & \bop n\cdot m)\bcp_j\cr
        (,\bop n)\bcp & n_i\cr
    }}
}
\medskip

\noindent $(n+m)_j$ means $n+m$ with a priority of $j$, not $\bop n+m\bcp_j$.
So for example expanding $2\cdot((1+2)\cdot2)+1$,
$$ \eqalign{
    2_\infty*_2{(_\infty}{(_\infty}1_\infty+_12_\infty{)_0}*_22_\infty{)_0}+_11_\infty &\longto \bop2*\bcp_2{(_\infty}{(_\infty}1_\infty+_12_\infty{)_0}*_22_\infty{)_0}+_11_\infty\cr
    &\longto \bop2*\bcp_2{(_\infty}{(_\infty}\bop1+\bcp_12_\infty{)_0}*_22_\infty{)_0}+_11_\infty\cr
    &\longto \bop2*\bcp_2{(_\infty}{(_\infty}\bop1+\bcp_1\bop2)\bcp_0*_22_\infty{)_0}+_11_\infty\cr
    &\longto \bop2*\bcp_2{(_\infty}{(_\infty}\bop3)\bcp_0*_22_\infty{)_0}+_11_\infty\cr
    &\longto \bop2*\bcp_2{(_\infty}3_\infty*_22_\infty{)_0}+_11_\infty\cr
    &\longto \bop2*\bcp_2{(_\infty}\bop3*\bcp_22_\infty{)_0}+_11_\infty\cr
    &\longto \bop2*\bcp_2{(_\infty}\bop3*\bcp_2\bop2)\bcp_0+_11_\infty\cr
    &\longto \bop2*\bcp_2{(_\infty}\bop6)\bcp_0+_11_\infty\cr
    &\longto \bop2*\bcp_26_\infty+_11_\infty\cr
    &\longto \bop2*\bcp_2\bop6+\bcp_11_\infty\cr
    &\longto \bop12+\bcp_11_\infty\cr
    &\longto \bop12+\bcp_11_0\cr
    &\longto 13_0\cr
} $$

Notice that in both examples, the characters in the string were given an initial priority.
Thus we can extend our definition of an expander to include an {\it initial priority} function $\pi\colon\Sigma\plongto\overline{\bb N}$ which gives characters their initial priority.
We can then canonically extend this to a function $\pi\colon\Sigma^*\plongto(\Sigma\times\overline{\bb N})^*$ defined by $\pi(\sigma^1\cdots\sigma^n)=\sigma^1_{\pi(\sigma^1)}\cdots\sigma^n_{\pi(\sigma^n)}$.
Then a $\beta$-reduction of a string $\xi\in\Sigma^*$ is taken to mean a $\beta$-reduction of $\pi(\xi)$.

Notice that once again we require that $\pi$ only be a partial function.
This is since that we don't always need every character in $\Sigma$ to have an initial priority; in the above examples there's no need to give $\bop n+\bcp$ an inital priority, its priority is given through the $\beta$-reduction of $n$ and $+$.
So we now provide a new definition of a {\it program}, which is a string $\xi=\sigma^1\cdots\sigma^n\in\Sigma^*$ such that $\pi(\sigma^i)$ exists for all $1\leq i\leq n$.
We can only of course discuss the reductions of programs, as $\pi(\xi)$ is only defined if $\xi$ is a program.

\subsection{Stated Expansion}

Suppose we'd like to expand a program with variables in it.
Then we cannot just use the previous definitions, as the actions of $\sigma$ (which is to be understood as the function $\beta(\sigma,\bullet)$) are determined before any expansion occurs.
We need a way to store the value of variables, a state.

This leads us to the following definition: let $\Sigma_P$ and $\Sigma_A$ be two disjoint sets of symbols: $\Sigma_P$ the set of {\it printable symbols} and $\Sigma_A$ the set of {\it abstract symbols}.
$\Sigma_P$ will generally be a set consisting of the string representations of abstract symbols, be it operators like $+$ and $\cdot$ or variable names.
$\Sigma_A$ are the actual objects which can ``execute something''.
Let us further define $\Sigma=\Sigma_P\dcup\Sigma_P$ and let $\State=\set{\Sigma_P\longto\Sigma_A\dcup\set\epsilon}$ be the set of all functions from printable to abstract symbols.
We allow printable symbols to be mapped to $\epsilon$, the empty word; this should be thought of the symbol having no value.

Now we begin with an initial $\beta$ function which is a partial function
$$ \beta\colon\overline\Sigma_A\times\overline\Sigma\times\State \plongto \bigl(\overline\Sigma\dcup\set\epsilon\bigr)\times\State $$
Recall that $\overline\Sigma$ is $\Sigma\times{\bb N}$.
We will denote tuples in $X\times\State$ by $\gen{x,s}$ for $x\in X$ and $s\in\State$ for the sake of readability.
So we now wish to extend to a $\beta$ function
$$ \beta\colon\overline\Sigma^*\times\State \plongto \overline\Sigma^*\times\State $$
We do this as follows: given $\xi\in(\Sigma\times\overline{\bb N})^*$ and $s\in\State$ we define $\beta\gen{\xi,s}$ as follows:
\benum
    \item if $\xi=\sigma_i\xi'$ for $\sigma\in\Sigma_P$ then $\beta\gen{\xi,s}=\gen{s(\sigma)_i\xi',s}$,
    \item if $\xi=\sigma^1_i\sigma^2_j\xi'$ for $\sigma^1\in\Sigma_A$, $i\geq j$, such that $\beta\gen{\sigma^1_i\sigma^2_j,s}=\gen{\sigma^3_k,s'}$ is defined, then
        $\beta\gen{\xi,s}=\gen{\sigma^3_k\xi',s'}$,
    \item otherwise if $\beta\gen{\sigma^2_j\xi',s}=\gen{\xi'',s'}$ then $\beta\gen{\xi,s}=\gen{\sigma^1_i\xi'',s'}$.
\eenum
We also define the {\it initial priority function} to be a map $\pi\colon\Sigma_P\longto\overline{\bb N}$ (this is not a partial function: every printable symbol must be given a priority).
This is once again canonically extended to a function $\pi\colon\Sigma_P^*\longto(\Sigma_P\times\overline{\bb N})^*$.
And an {\it initial state} $s_0\in\State$.
The quintuple $(\Sigma_P,\Sigma_A,\beta,\pi,s_0)$ is called an {\it expander}.
The expansion of a string $\xi\in S$ is the process of iteratively applying $\beta$ to $\gen{\pi(\xi),s_0}$.

\medskip
{\bf Example:\/} So for example let
$$ \eqalign{
    \Sigma_P &= {\bb N}\cup\set{+,\cdot,=,;}\cup\set{{\tt let}}\cup\set{x^i}[i\in{\bb N}],\cr
    \Sigma_A &= {\bb N}\cup\set{(n+),(n\cdot)}[n\in{\bb N}]\cup\set{{\tt let}}\cup\set{({\tt let}x^i),({\tt let}x^i=)}[i\in{\bb N}]
} $$
where the natural numbers in $\Sigma_A$ are not the same as the natural numbers in $\Sigma_P$ since they must be disjoint, same for {\tt let}.
But they both essentially represent the same thing: $s_0$ maps $n\mapsto n$ for $n\in{\bb N}$ (the left-hand $n$ is in $\Sigma_p$, the right-hand $n$ is in $\Sigma_A$) and ${\tt let}\mapsto{\tt let}$.
All other printable symbols are mapped to $\epsilon$.

And similar to the previous example we define $\pi(n)=\infty$, $\pi(+)=1$, and $\pi(\cdot)=2$.
We extend this to $\pi(;)=0$, $\pi(=)=0$, $\pi({\tt let})=\infty$, and $\pi(x^i)=\infty$.

Let us take the same transitions as the example in the previous section for $n,(n+),(n\cdot)$ (we have to add the condition that the state doesnt change).
We further add the transitions
\medskip
\centerline{
    \vtop{\ialign{\hfil$#$\hfil\tabskip=.5cm&\hfil$#$\hfil\cr
        \gen{\sigma^1\sigma^2,s} & \beta\gen{\sigma^1\sigma^2,s}\cr\noalign{\kern3pt\hrule\kern3pt}
        \gen{\sigma;,s} & \gen{\sigma_0,s}\cr
        \gen{\lett x^i,s} & \gen{(\lett x^i)_\infty,s}\cr
        \gen{(\lett x^i)=,s} & \gen{(\lett x^i=)_0,s}\cr
        \gen{(\lett x^i=)n,s} & \gen{\epsilon,s[x^i\mapsto s(n)]}\cr
}}}

\noindent In the final transition, $n\in\Sigma_A$.
Then for example (we will be skipping trivial expansions):
$$ \eqalign{
    \lett x^1=1+2;\ \lett x^2=2;\ x^1\cdot x^2; &\longto
    {\tt let}_\infty x^1_\infty=_01_\infty+_12_\infty;_0\ {\tt let}_\infty x^2_\infty =_0 2_\infty;_0\ x^1_\infty\cdot_2 x^2_\infty;_0\cr
    s_0\quad&\longto ({\tt let}x^1=)_01_\infty+_12_\infty;_0\ {\tt let}_\infty x^2_\infty =_0 2_\infty;_0\ x^1_\infty\cdot_2 x^2_\infty;_0\cr
    s_0\quad&\longto ({\tt let}x^1=)_03_0\ {\tt let}_\infty x^2_\infty =_0 2_\infty;_0\ x^1_\infty\cdot_2 x^2_\infty;_0\cr
    s_0[x^1\mapsto3]\quad&\longto {\tt let}_\infty x^2_\infty =_0 2_\infty;_0\ x^1_\infty\cdot_2 x^2_\infty;_0\cr
    s_0[x^1\mapsto3,\,x^2\mapsto2]\quad&\longto x^1_\infty\cdot_2 x^2_\infty;_0\cr
    s_0[x^1\mapsto3,\,x^2\mapsto2]\quad&\longto 3_\infty\cdot_2 x^2_\infty;_0\cr
    s_0[x^1\mapsto3,\,x^2\mapsto2]\quad&\longto (3\cdot)_2 x^2_\infty;_0\cr
    s_0[x^1\mapsto3,\,x^2\mapsto2]\quad&\longto (3\cdot)_2 2_\infty;_0\cr
    s_0[x^1\mapsto3,\,x^2\mapsto2]\quad&\longto 6_0\cr
} $$

\bye

